---
title: "Pathway-protein correlations constrained to DA pathways and DA proteins"
output: html_notebook
author: Zakhar Krekhno
---

```{r, echo=FALSE, include=FALSE}
library(readxl)
library(reshape2)
library(psych)
library(tidyverse)
theme_set(theme_classic(base_size = 14))
```

Import all the necessary matrices for analysis
```{r import}
files <- list.files(path = here::here("corr_mats/"))
pth_file_list <- str_subset(files,"clr(.)+pth_da")
ms_file_list <- str_subset(files, "ms")
#Import pathway counts
pth_list <- map(pth_file_list, 
                    ~read_csv(file = here::here(str_c("corr_mats/",.))))
pth_mat_list <- pth_list %>% 
  map(column_to_rownames, var = "...1") %>% 
  map(as.matrix)
#Import MS data
mq_mat <- read_csv(file = here::here(str_c("corr_mats/",ms_file_list))) %>% 
  column_to_rownames(var = "...1") %>% 
  as.matrix()
```

Import the metadata and create sub-divided matrices.
We will be doing correlations within just Ileum or Colon and within noIBD, active, or inactive IBD
```{r import-meta}
meta_full <- read_excel(here::here("Metadata.xlsx")) %>% 
  select("sample" = `#SampleID`, "group" = Group, everything()) %>% 
  mutate(sample = str_remove(sample, "Rep")) %>% 
  filter(sample %in% rownames(mq_mat)) %>% 
  mutate(across(matches("(Active)|(TI_resection)|(Cholestyramine)"), ~as.character(fct_recode(as_factor(.x), 
                                                Yes = "1", No = "0", Not_applicable = "NA"))))
#Prepare metadata tables
meta_group_list <- vector(mode = "list", length = 4) %>% 
  map(~ meta_full) %>%  
  map2(rep(c("Colon","Ileum"), each = 2), 
       ~filter(.x, Tissue_type == .y)) %>% 
  map2(rep(c("Control","IBD"),times = 2), 
       ~filter(.x, group == .y)) %>% 
  set_names(nm = str_c(rep(c("Colon","Ileum"), each = 2), rep(c("Control","IBD"),times = 2), sep = "_"))
meta_active_list <- vector(mode = "list", length = 6) %>% 
  map(~ meta_full) %>%  
  map2(rep(c("Colon","Ileum"), each = 3), 
       ~filter(.x, Tissue_type == .y)) %>% 
  map2(rep(c("Not_applicable","No","Yes"),times = 2), 
       ~filter(.x, Active_IBD == .y)) %>% 
  set_names(str_c(rep(c("Colon","Ileum"), each = 3),rep(c("Not_applicable","No","Yes"),times = 2), sep = "_"))
meta_list <- c(meta_group_list, meta_active_list)

#Keep list names for convenience
subgroups <- names(meta_list)
```

Make appropriate matrices for each group
```{r subset_pth}
pth_filt_list <- meta_list %>% 
  map(~pth_mat_list [[1]]) %>% 
  map2(meta_list, ~ .x [pull(.y,var = sample),]) 
```

Further filter the proteomics data to just include DE proteins between active/inactive IBD vs control in each group
```{r select_top_prots}
#Import the results
da_res <- map(list.files(here::here("spear_subset_on_Foster"), pattern = "full_"),
              ~read_csv( file = here::here(str_c("spear_subset_on_Foster/",.)))) 
da_res <- set_names(da_res, nm = list.files(here::here("spear_subset_on_Foster"), pattern = "full_"))
#Reduce the results to one table, and filter afterwards
da_res_interest <- da_res %>% 
  reduce(left_join, by = c("gene.name","first.id")) %>% 
  select(gene.name,first.id, contains("padj"), -contains("grand")) %>% 
  filter(if_any(contains("adj"), ~.<0.01))
#Get the protein names to select
da_prots <- da_res_interest %>% 
  select(gene.name) %>% 
  separate(gene.name, into = letters [1:20], sep = " ") %>% 
  pivot_longer(letters [2:20], names_to = "letters", values_to = "secondary_names")
da_prots_final <- c(da_prots$a,da_prots$secondary_names) %>% 
  . [!is.na(.)] %>% 
  str_remove_all(";")

#Create the new mq_list
mq_top_list <- map(meta_list, ~ mq_mat [,colnames(mq_mat) %in% da_prots_final]) %>% 
  map2(meta_list, ~ .x [pull(.y,var = sample),]) 
#Ensure rows are in the same order for the matrices
all(map2_lgl(pth_filt_list, mq_top_list, ~all(rownames(.x) == rownames(.y))))
```

Filter OTUs to only include Enterobacteriacea

Run correlations on subset proteins

```{r spearman_top_da, warning=FALSE, eval=FALSE}
spear_top_list <- map2(mq_top_list,pth_filt_list,
                     ~corr.test(x = .x, y = .y, use = "pairwise", method="spearman", adjust="BH", ci=FALSE))
coef_top_list <- map2(rep(spear_top_list,each=3),rep(c("r","p","p.adj"),times = length(spear_top_list)), ~melt(.x [.y]))

rs_top_list <- coef_top_list [seq(from = 1, to = length(coef_top_list), by = 3)]
ps_top_list <- coef_top_list [seq(from = 2, to = length(coef_top_list), by = 3)]
padj_top_list <- coef_top_list [seq(from = 3, to = length(coef_top_list), by = 3)]

df_top_list <- pmap(list(rs_top_list,ps_top_list,padj_top_list),
               ~reduce(list(..1,..2,..3),left_join, by = c("Var1","Var2")))
names(df_top_list) <- names(meta_list)
#Filter the results to only include somewhat meaningful work
df_top_filt_list <- df_top_list %>% 
  map(~filter(., value < 0.05))
df_top_cors <- df_top_filt_list %>% 
   map(~arrange(.,value,desc(abs(as.numeric(value.x))))) %>% 
  map(slice_head,n=1) %>% 
  keep(~nrow(.)>0) %>% 
  map(mutate, Var1 = as.character(Var1), Var2 = as.character(Var2))
#Print out correlation results
#walk2(df_top_list, names(ds_top_list),
 #     ~write_csv(.x,here::here(str_c("spear_pathways/",.y,"_pth.csv"))))

```

Plot scatter plots of top correlations

```{r scatter-top}
#Select just protein and otu as selected top correlation
mq_select_list <- mq_top_list %>% 
  keep(names(.) %in% names(df_top_cors)) %>% 
  map2(df_top_cors, ~ .x [,colnames(.x) %in% pull(.y,var = "Var1")]) %>%  
  map(as.data.frame, stringsAsFactors = F)
pth_select_list <- pth_filt_list %>% 
  keep(names(.) %in% names(df_top_cors)) %>%  
  map2(df_top_cors, ~ .x [,colnames(.x) %in% pull(.y,var = "Var2")]) %>%
  map(as.data.frame, stringsAsFactors = F)
#prepare list for scatter plots
scatter_list <- map2(mq_select_list, pth_select_list, merge, by = 0)
#prepare lists for names
xlabs <- map_chr(df_top_cors, pull, var = "Var1")
ylabs <- map_chr(df_top_cors, pull, var = "Var2")
rcoefs <- map_chr(df_top_cors, pull, var = "value.x")
pvals <- map_chr(df_top_cors, pull, var = "value")

#pdf(here::here("spear_pathways/Scatter plots of top correlations for prevalent DA pathways.pdf"))
pwalk(list(scatter_list, xlabs,ylabs, rcoefs,pvals, names(scatter_list)), 
      ~print(ggplot(..1, aes(x = `.x[[i]].x`, y = `.x[[i]].y`)) +
               geom_point() +
               geom_smooth(method = "lm") + 
               labs(x = ..2, y = ..3, 
                    subtitle = str_c("Spearman rho =",round(as.numeric(..4),2)," Adjusted p=",
                                     format(as.numeric(..5),digits = 2, scientific = T)), title = ..6)
             ))
  
#dev.off()
```
Plot bubble plots of Spearman correlations
```{r spearman-plot-bubble, fig.height=6, fig.width=8}
#Only keep strong associations
df_top_reduced <- keep(df_top_filt_list, ~nrow(.) > 0) %>% 
  map2(c(0.51,0.32,0.2),~filter(.x,abs(value.x) > .y, value < 0.05))



#pdf(here::here("spear_pathways/Prevalent DA Pathway correlations.pdf"), height = 8, width = 10)
walk2(df_top_reduced, names(df_top_reduced),
      ~print(ggplot(.x, aes_(x=quote(Var1), y=quote(Var2))) +
              geom_point(aes(color = value.x, fill = value.x, size = value), alpha = 0.6, shape = 21) +
              scale_fill_distiller(name = "Spearman rho",palette = "PuOr", breaks = c(-0.4,0,0.4)) +
              scale_color_distiller(name = "Spearman rho", palette = "PuOr",breaks = c(-0.4,0,0.4)) +
              scale_size(name = "Adjusted p",trans = 'reverse', range = c(5,8)) +
              labs(x=element_blank(), y=element_blank(),
                   title = .y) +
              theme_bw(base_size = 12) + theme(axis.text.x = element_text(angle = 45, hjust = 0.9))))
#dev.off()
```
Replot graphs with just proteins that were correlated with Entero
```{r entero-prot-constrain-plot, fig.width=10, fig.height=10}
entero_assoc_files <- list.files(path = here::here("spear_entero/"), pattern = ".csv")
entero_assoc_df <- map(entero_assoc_files, ~read_csv(here::here(str_c("spear_entero/",.)))) %>% 
  set_names(str_remove(entero_assoc_files,".csv"))
entero_assoc_signi <- entero_assoc_df %>% 
  map(filter, value < 0.05) %>% 
  keep(~nrow(.) >0) %>% 
  map(pull, "Var1") %>% 
  reduce(append) %>% 
  unique()
#Constrain correlations to Entero-associated proteins
df_top_entero <- keep(df_top_filt_list, ~nrow(.) > 0) %>% 
  map(filter, Var1 %in% entero_assoc_signi) %>% 
  keep( ~nrow(.) > 0)
#pdf(here::here("spear_pathways/DA Pathway correlations constrained to Entero-associated proteins.pdf"), height = 10, width = 10)
walk2(df_top_entero, names(df_top_entero),
      ~print(ggplot(.x, aes_(x=quote(Var1), y=quote(Var2))) +
              geom_point(aes(color = value.x, fill = value.x, size = value), alpha = 0.6, shape = 21) +
              scale_fill_distiller(name = "Spearman rho",palette = "PuOr", breaks = c(-0.4,0,0.4)) +
              scale_color_distiller(name = "Spearman rho", palette = "PuOr",breaks = c(-0.4,0,0.4)) +
              scale_size(name = "Adjusted p",trans = 'reverse', range = c(5,8)) +
              labs(x=element_blank(), y=element_blank(),
                   title = .y, subtitle = "Entero-associated DA proteins") +
              theme_bw(base_size = 10) + theme(axis.text.x = element_text(angle = 45, hjust = 0.9))))
#dev.off()
#walk2(df_top_entero, names(df_top_entero), 
 #     ~write_csv(.x, here::here(str_c("spear_pathways/",.y,"_DA_pathway_Entero_associated_prot.csv"))))
```


